---
title: 'Reproducible Data: Peer Assessment 1'
author: "Robert Wiederstein"
date: "June 1, 2015"
output: html_document
---
First, the data set will be read into R.

```{r, echo=TRUE}
file <- "activity.zip"
path <- getwd()
zipfile <- paste(path, file, sep = "/")
unzip(zipfile = zipfile)
a <- read.csv("activity.csv", stringsAsFactors = F)
```



```{r, echo=TRUE}
library(plyr)
library(dplyr)
a <- a[which(complete.cases(a) == TRUE), ]
tot.steps.per.day <- ddply(a, .(date), summarize,
                           date = unique(date),
                           sum = sum(steps, na.rm = T), 
                           mean = mean(steps, na.rm = T), 
                           median = median(steps, na.rm = T)
                           )
```
The mean total number of steps taken per day is `r round(tot.steps.per.day$mean, 2)`.

```{r}
library(ggplot2)
p <- ggplot(tot.steps.per.day, aes(sum))
p <- p + geom_histogram()
p <- p + ggtitle("Histogram of Total Steps by Day")
p
```


```{r}
a <- a[which(complete.cases(a) == T), ]
a[which(nchar(a$interval) == 3), 3] <- paste("0", a[which(nchar(a$interval) == 3), 3], sep = "")
a[which(nchar(a$interval) == 2), 3] <- paste("00", a[which(nchar(a$interval) == 2), 3], sep = "")
a[which(nchar(a$interval) == 1), 3] <- paste("000", a[which(nchar(a$interval) == 1), 3], sep = "")
a$steps <- as.integer(a$steps)
avg.steps.per.interval <- ddply(a, .(interval), summarize, 
                                mean_steps = round(mean(steps, na.rm = T), 3),
                                time_interval = unique(interval)
                                )
avg.steps.per.interval$time_interval <- strptime(avg.steps.per.interval$time_interval, "%H%M")
avg.steps.per.interval$hm <- format(avg.steps.per.interval$time_interval, format = "%H:%M")
avg.steps.per.interval$hm <- as.POSIXct(avg.steps.per.interval$hm, format = "%H:%M")

library(scales)
library(ggplot2)
p <- ggplot(avg.steps.per.interval, aes(hm, mean_steps))
p <- p + geom_line()
p <- p + xlab("time") + ylab("number") + ggtitle("Avg. Steps Per Time Interval \n (intervals in 5 min. increments)")
p <- p + scale_x_datetime(labels = date_format("%H%:%M"))
p
```


```{r, echo=T}
time.interval <- avg.steps.per.interval[which(avg.steps.per.interval$mean_steps == max(avg.steps.per.interval$mean_steps)), 1]
```

The 5-minute interval where the most steps are taken, on average, is `r time.interval`.


```{r, echo=TRUE}
b <- read.csv("activity.csv", stringsAsFactors = F)
missing.obs <- length(which(complete.cases(b) == F))
```

The number of observations that have missing or incomplete data are `r missing.obs`.

```{r, echo=T}
b <- read.csv("activity.csv", stringsAsFactors = F)
count.na <- function(x){
    length(which(is.na(x) == T))
}
missing.na <- apply(b, MARGIN = 2, count.na)
```

All of the missing values in the data set are confined to the variable "steps."

```{r, echo=TRUE}
complete <- b[which(complete.cases(b) == T), ]

complete.steps <- ddply(complete, .(interval), summarize,
                           date = unique(interval),
                           sum = sum(steps, na.rm = T), 
                           mean = mean(steps, na.rm = T), 
                           median = median(steps, na.rm = T)
                           )
#complete.steps <- complete.steps[, c(1, 4)]
missing <- b[which(complete.cases(b)==F), ]
impute <- merge(missing, complete.steps, by = c("interval"))
impute <- impute[order(impute$date.x), ]
impute$steps <- impute$mean
impute <- impute[, c("steps", "date.x", "interval")]
names(impute)[2] <- "date"
new.df <- rbind(complete, impute)
```

```{r, echo=TRUE}
imputed.steps <- ddply(new.df, .(interval), summarize,
                           date = unique(interval),
                           sum = sum(steps, na.rm = T), 
                           mean = mean(steps, na.rm = T), 
                           median = median(steps, na.rm = T)
                           )

p <- ggplot(imputed.steps, aes(sum))
p <- p + geom_histogram()
p <- p + ggtitle("Histogram Total Num. of Steps Each Day")
p
```
The mean steps from the imputed data set is `r imputed.steps$mean`.  The median steps from the imputed data set is `r imputed.steps$median`.

```{r, echo=TRUE}
p <- ggplot(imputed.steps, aes(date, sum))
p <- p + geom_line(colour = "red")
p <- p + geom_line(data = complete.steps)
p <- p + ggtitle("Compare Total Steps in Imputed Dataset to Complete Cases Dataset")
p
```

```{r, echo=TRUE}
new.df$weekdays <- weekdays(as.Date(new.df$date), abbreviate = T)
weekdays <- c("Mon", "Tue", "Wed", "Thu", "Fri")
weekend <- c("Sat", "Sun")
new.df$weekdays[which(new.df$weekdays %in% weekdays)] <- "weekday"
new.df$weekdays[which(new.df$weekdays %in% weekend)] <- "weekend"
new.df$weekday <- as.factor(new.df$weekday)

new.df1 <- ddply(new.df, .(interval, weekday), summarize,
                           interval = unique(interval),
                           sum = sum(steps, na.rm = T), 
                           mean = mean(steps, na.rm = T), 
                           median = median(steps, na.rm = T)
                           )

p <- ggplot(new.df1, aes(interval, sum))
p <- p + geom_line()
p <- p + facet_grid(weekday ~ .)
p <- p + ggtitle("Average Number of Steps Taken \n Weekend vs. Weekday")
p <- p + ylab("Number of Steps")
p
```